---
title: "Exam 1 - Part 2: Statistical Analysis"
subtitle: "AGST 50104 - Experimental Design"
author: "Sam Fernandes"
format:
  pdf:
    toc: false
    colorlinks: true
    geometry:
      - top=1in
      - bottom=1in
      - left=1in
      - right=1in
editor: source
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = TRUE,
    message = FALSE,
    warning = FALSE,
    comment = "#>",
    fig.width = 7,
    fig.height = 5
)

# Load required packages
library(tidyverse)
library(car) # For Type III SS via Anova()
library(emmeans) # For post-hoc comparisons
library(broom) # For tidy model outputs
library(MASS) # For Box-Cox transformation
library(scales) # For enhanced statistical plots
```

# Question 1

The following data refer to an experiment comparing varieties with respect to cotton properties used in fiber manufacturing. The characteristic selected for analysis was the Pressley strength index. This index was determined by measuring the breaking tension of a bundle of fibers with a pre-fixed cross-sectional area. The experiment was conducted using a randomized complete block design with three replications. The replications consisted of three strength-testing machines, each operated by a different technician. What can you conclude after proceeding with the proper analysis for this experiment? Use figures to illustrate your conclusions.

| Cultivar | Machine I | Machine II | Machine III | Mean  |
|----------|-----------|------------|-------------|-------|
| V1       | 7.62      | 8.00       | 7.93        | 7.85  |
| V2       | 8.14      | 8.15       | 7.87        | 8.05  |
| V3       | 7.76      | 7.73       | 7.74        | 7.74  |
| V4       | 7.17      | 7.57       | 7.80        | 7.51  |
| V5       | 7.46      | 7.68       | 7.21        | 7.45  |
| V6       | 7.30      | 7.70       | 7.50        | 7.50  |

## Answer

Model (RCBD):

$$
y_{ij} = \mu + \tau_i + \beta_j + \varepsilon_{ij}, \quad \varepsilon_{ij} \sim N(0,\sigma^2)
$$

where $\tau_i$ is cultivar effect and $\beta_j$ is machine (block) effect.

```{r q1-data-model}
q1_data <- tibble::tribble(
    ~cultivar, ~machine, ~strength,
    "V1", "I", 7.62,
    "V1", "II", 8.00,
    "V1", "III", 7.93,
    "V2", "I", 8.14,
    "V2", "II", 8.15,
    "V2", "III", 7.87,
    "V3", "I", 7.76,
    "V3", "II", 7.73,
    "V3", "III", 7.74,
    "V4", "I", 7.17,
    "V4", "II", 7.57,
    "V4", "III", 7.80,
    "V5", "I", 7.46,
    "V5", "II", 7.68,
    "V5", "III", 7.21,
    "V6", "I", 7.30,
    "V6", "II", 7.70,
    "V6", "III", 7.50
) |>
    mutate(cultivar = factor(cultivar), machine = factor(machine))

q1_model <- lm(strength ~ machine + cultivar, data = q1_data)
```

### Assumptions

```{r q1-assumptions}
q1_shapiro <- shapiro.test(residuals(q1_model))
q1_levene <- leveneTest(strength ~ cultivar, data = q1_data)

q1_shapiro
q1_levene
```

Interpretation:

- Normality is acceptable (Shapiro-Wilk $p = 0.8975$).
- Homogeneity of variance is acceptable (Levene $p = 0.659$).

### ANOVA and mean comparison

```{r q1-anova-emmeans}
q1_anova <- Anova(q1_model, type = 3)
q1_anova

q1_emm <- emmeans(q1_model, ~cultivar)
q1_emm

pairs(q1_emm, adjust = "tukey")
```

Interpretation:

- Machine effect is not significant ($p = 0.164$).
- Cultivar effect is significant ($p = 0.0186$).
- V2 has the highest adjusted mean (about $8.05$), and Tukey results show V2 is significantly higher than V5 and V6 at $\alpha = 0.05$.

### Figure

```{r q1-figure}
ggplot(q1_data, aes(x = cultivar, y = strength, fill = cultivar)) +
    geom_boxplot(alpha = 0.6, outlier.shape = NA) +
    geom_point(
        aes(color = machine),
        position = position_jitter(width = 0.08),
        size = 2.4,
        alpha = 0.85
    ) +
    labs(
        title = "Pressley Strength by Cultivar (RCBD)",
        x = "Cultivar",
        y = "Pressley Strength Index",
        color = "Machine",
        fill = "Cultivar"
    ) +
    theme_minimal(base_size = 12)
```

Final conclusion for Question 1:

The RCBD analysis indicates cultivar differences are present. V2 is the best performer, while machine-to-machine variability was not statistically important in this experiment.

\clearpage

# Question 2

The data set bacteria.csv is from an experiment that evaluated bacterial growth on 24 treatments. It is a Completely Randomized Design (CRD) with four replications.

a. Formulate the statistical hypotheses $H_0$ and $H_1$ related to the treatments.
b. Using appropriate tests, verify the basic assumptions of the statistical model for performing ANOVA. Interpret the results.
c. In the event of violation of any assumptions, assess the possibility of determining an appropriate transformation using Box-Cox (1946). Then, recheck assumptions.
d. Perform ANOVA on the appropriate scale at 5% significance. Present and interpret results. Identify treatment with less bacterial growth.

## Answer

### (a) Hypotheses

Let $\mu_i$ be the mean bacterial growth for treatment $i$, $i=1,\ldots,24$.

$$
H_0: \mu_1 = \mu_2 = \cdots = \mu_{24}
$$

$$
H_1: \text{at least one treatment mean differs}
$$

### Data preparation

```{r q2-data}
q2_data <- read.csv("bacteria.csv") |>
    pivot_longer(cols = starts_with("r"), names_to = "rep", values_to = "growth") |>
    mutate(treat = factor(treat), rep = factor(rep))

head(q2_data)
```

### (b) Assumptions on raw scale

```{r q2-raw-assumptions}
q2_model_raw <- lm(growth ~ treat, data = q2_data)

q2_raw_shapiro <- shapiro.test(residuals(q2_model_raw))
q2_raw_levene <- leveneTest(growth ~ treat, data = q2_data)

q2_raw_shapiro
q2_raw_levene
```

Interpretation:

- Normality is violated on raw scale (Shapiro-Wilk $p = 0.000123$).
- Homogeneity is also violated on raw scale (Levene $p = 0.000353$).

### (c) Box-Cox transformation and re-check

```{r q2-boxcox-transform}
q2_boxcox <- boxcox(q2_model_raw, plotit = FALSE)
q2_lambda <- q2_boxcox$x[which.max(q2_boxcox$y)]
q2_lambda

q2_data <- q2_data |>
    mutate(growth_bc = (growth^q2_lambda - 1) / q2_lambda)

q2_model_bc <- lm(growth_bc ~ treat, data = q2_data)

q2_bc_shapiro <- shapiro.test(residuals(q2_model_bc))
q2_bc_levene <- leveneTest(growth_bc ~ treat, data = q2_data)

q2_bc_shapiro
q2_bc_levene
```

Interpretation:

- Box-Cox suggests $\lambda \approx 0.2$.
- After transformation, assumptions are acceptable:
  - Shapiro-Wilk $p = 0.395$
  - Levene $p = 0.0846$

### (d) ANOVA on appropriate scale

```{r q2-anova}
q2_anova <- anova(q2_model_bc)
broom::tidy(q2_anova)
```

Interpretation:

- Treatment effect is highly significant ($p < 0.001$; approximately $8.62 \times 10^{-42}$).
- Reject $H_0$. Not all treatment means are equal.

### Treatment with less growth

```{r q2-lowest-treatment}
q2_means <- q2_data |>
    group_by(treat) |>
    summarize(
        mean_raw = mean(growth),
        mean_bc = mean(growth_bc),
        .groups = "drop"
    ) |>
    arrange(mean_raw)

q2_means
```

Interpretation:

- The treatment with less bacterial growth is treatment 20 (raw mean $= 12.75$).

```{r q2-anova-emmeans}
q2_emm <- emmeans(q2_model_bc, ~treat)
q2_emm

model_means_cld <- multcomp::cld(q2_emm,  adjust = "Tukey", Letters = letters, alpha = 0.05)
model_means_cld
```



### Figure

```{r q2-figure}
ggplot(q2_data, aes(x = treat, y = growth)) +
    geom_boxplot(fill = "#80c5c5", alpha = 0.65, outlier.alpha = 0.45) +
    labs(
        title = "Bacterial Growth by Treatment (CRD)",
        x = "Treatment",
        y = "Bacterial Growth"
    ) +
    theme_minimal(base_size = 11) +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

Final conclusion for Question 2:

Raw assumptions failed, Box-Cox transformation corrected assumptions, ANOVA confirmed strong treatment differences, and treatment 20 showed the lowest bacterial growth.

